The Rockerverse: Images, Packages, and Applications for Containerization with R

### 2020-01-22 manuscript.Rmd
Added repo2docker lessons learned

**rocker-win** ([https://github.com/nuest/rocker-win](https://github.com/nuest/rocker-win)) is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
These base images match different Windows versions: Windows Server 2019, Windows Server 2016, and Windows Server, version 1803.
The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker&nbsp;Hub](https://hub.docker.com/r/nuest/rocker-win).
The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
All these images can be run on Docker for Desktop on Windows 10.
New R versions are only added on demand.
The explorative rocker-win project demonstrates the variety of Windows on Docker, which is potentially confusing, e.g., because of the different base images.
The proof of concept provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \CRANpkg{plumber}, \CRANpkg{Shiny}, or R packages with system dependencies such as \CRANpkg{sf}.
The latter is quite comfortable and fast actually, compared to the default installation from source on Linux, because CRAN ships pre-compiled binaries for Windows.
However, compared to the various Linux images, the footprint of Windows images is quite large.

For Windows, just as originally for Linux, the cloud use cases drive the development and Docker and Microsoft collaborate closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
Naturally, a Windows license is required for the host machine, and the licensing impacts the potential to build images in clouds or redistribute exported images as archive files.
But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.

## Non-Rocker Linux images
\label{nondebian}

\label{rhub}
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \CRANpkg{rhub} \citep{csardi_rhub_2019}.
The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
The `Dockerfile`s and images are published on Docker&nbsp;Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\nameref{development}) not for reproducible workflows.
_The rationale_ of not re-using the Rocker image stack is the high number of operatings systems and configurations, which can be more uniformly maintained in a seperate suite of images.
Also, a common structure across the respective image stacks of the operating systems, e.g., for `debian`, adding GCC in `debian-gcc` and then R's development version in `debian-gcc-devel`.
The platform also has a specific user configuration differing from the setup covering typical applications of Rocker images, and the images include considerably more software, such as a fairly complete LaTeX installation, to provide an experience closer to CRAN.

@@

_Data Science_ is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
The discussions shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
The practice of Data Science often involves a combination tools and software stacks and requires a cross-cutting skillset.
This complexity and an inherent concern for openness and reproducibility in the Data Science community lead to Docker being used widely.
<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
This section presents exemplary Docker images and image stacks featuring R intended for Data Science.

@@

**Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and also configures the \CRANpkg{reticulate} package.
The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.

@@

As a further example, [**Radiant** project](https://radiant-rstats.github.io/docs/) provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9GB.

@@

**Gigantum** ([http://gigantum.com/](http://gigantum.com/) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects.
It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
The independent image stack can be traced back to the Gigantum environment and its features.
The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.} so that appropriate GPU drivers can be enabled automatically when supported.
Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusions}).
Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.

@@

Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images and launching a containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
As such, the choice of which package is most useful depends on the use-case at hand, but also the users level of expertise.

```{r clientfunctionalities, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
functionalities <- tribble(
  ~id, ~Functionality,
  "dockerfile", "Generate a Dockerfile",
  "build", "Build an image",
  "execute", "Execute a container locally or remotely",
  "cloud", "Deploy or manage an instances in the cloud",
  "interact", "Interact with an instance (e.g., file transfer)",
  "images", "Manage storage of images",
  "othercontainers", "Supports Docker and Singularity",
  "api", "Direct access to Docker API instead of using the CLI",
  "install", "Installing Docker software"
)

@@

**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is not actively maintained anymore, but should be honorably mentioned as the first R package for managing Docker images and containers.
It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an ssh connection, and has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
The output of container executions are converted to appropriate R types.
The Docker CLI, while evolving quickly and not with great concern for introducing breaking changes, basic functionality is unchanged for a long time so `harbor::docker_run(image = "hello-world")` works like a charm.

@@

**\CRANpkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
It interfaces with the Docker daemon over the Docker HTTP API\footnote{\href{https://docs.docker.com/engine/api/latest/}{https://docs.docker.com/engine/api/latest/}} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
The package is the only one not using system calls to the `docker` CLI for managing images and containers and easily exposes connections to remote Docker daemons, which has to be configured on the Docker level otherwise.
Using the API gives access to more information and is system independent and likely more reliable than parsing command line output.
\CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/tree/master/inst/spec}{https://github.com/richfitz/stevedore/tree/master/inst/spec} and \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.

**\CRANpkg{AzureContainers}** is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), see Section&nbsp;\nameref{azurecontainers}.
While it is mainly intended for working with Azure, as a convenience feature it includes lightweight, cross-platform shells to Docker and Kubernetes (tools `kubectl` and `helm`).
These can be used to create and manage arbitrary Docker images and containers, as well as Kubernetes clusters on any platform or cloud service.

**\CRANpkg{babelwhale}** allows executing and interacting with containers, which can use either Docker or Singularity as a backend \citep{cannoodt_babelwhale_2019}.
The package provides a unified interface to interact with Docker and Singularity containers.
Users can, for example, execute a command inside a container, mount a volume or copy a file.

@@

**\CRANpkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) with Docker: https://rstudio.github.io/renv/articles/docker.html

@@

**\CRANpkg{liftr}** ([https://nanx.me/liftr/](https://nanx.me/liftr/)) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
However, the reproducibility of such content authoring environments is often limited to individual machines --- it is not easy to replicate the system environment (libraries, R versions, R packages) where the document was compiled.
This issue becomes even more serious when it comes to collaborative document authoring and creating large-scale document building services.
_liftr_ solves this reproducibility problem by bringing Docker to the game.
In essence, _liftr_ helps R Markdown users create and manage Docker containers for rendering the documents, thus make the computations utterly reproducible across machines and systems.

@@

Such rApache applications have been used at [rapporter.net](http://rapporter.net), transforming parametrized R Markdown-like documents into interactive web applications for reporting.
A similar design was implemented at [CARD.com](https://www.card.com/), where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and rApache running in Docker containers:
the R session behind Rserve acted as a read-only, extremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
Similar dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker).
The images extend `rocker/r-base`, e.g., with an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/).
The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution.
For the use case at hand, the priority was to have more control over upstream updates, and to minimize the risk of breaking the automated builds of the Docker images and production jobs.

A more recent related work is the \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.

@@

\label{rocker-gpu}
Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible and useful \citep{haydel_enhancing_2015}.
GPUs are increasingly popular for compute-intensive machine learning (ML) tasks, e.g., deep artificial neural networks \citep{schmidhuber_deep_2015}.
Though in this case, containers are not completely portable between hardware environments, but the software stack for ML with GPUs is so complex to set up that a ready-to-use container is helpful.
Containers running GPU software require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugin for Docker and a set of base images with appropriate drivers and libraries.
The Rocker&nbsp;Project [has a repository](https://github.com/rocker-org/ml) with (beta) images based on these that include GPU-enabled versions of machine-learning R packages, e.g., `rocker/ml` and `rocker/tensorflow-gpu`.

@@

# Conclusions
\label{conclusions}

>>>>

**rocker-win** ([https://github.com/nuest/rocker-win](https://github.com/nuest/rocker-win)) is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
These base images match different Windows versions: Windows Server 2019, Windows Server 2016, and Windows Server, version 1803.
The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker&nbsp;Hub](https://hub.docker.com/r/nuest/rocker-win).
The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
All these images can be run on Docker for Desktop on Windows 10.
New R versions are only added on demand.
The explorative rocker-win project demonstrates the variety and pontentially confusing complexity of Windows on Docker, e.g., because of the different base images, but also provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \pkg{plumber\, \pkg{Shiny}, or R packages with system dependencies such as \pkg{sf}.
The latter is quite comfortable and fast actually, compared to installation from source on Linux, as R ships pre-compiled binaries for Windows.
However, compared to the various Linux images, the footprint of Windows images is quite large.

For Windows, just as originally for Linux, the cloud use cases drive the development and Docker and Microsoft collaborate closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
Naturally, a Windows license is required for the host machine, and the licensing impacts the potnetial to build images in clouds or redistribute exported images as archive files.
But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.

## Non-Debian Linux images
\label{nondebian}

\label{rhub}
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \pkg{rhub} \citep{csardi_rhub_2019}.
The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
The `Dockerfile`s and images are published on Docker&nbsp;Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\nameref{development}) not for reproducible workflows.
_The rationale_ of not re-using the Rocker image stack is the high number of operatings systems and configurations, which can be more uniformly maintained in a seperate suite of images.
Also, a common structure across the respective image stacks of the operating systems, e.g., for `debian`, adding GCC in `debian-gcc` and then R's development version in `debian-gcc-devel`.
The platform also has a specific user configuration differing from the setup covering typical applications of Rocker images, and the images include considerably more software, such as a fairly complete LaTeX installation, to provide an experience closer to CRAN.

@@

_Data Science_ is a widely discussed topic among all academic disciplines \citep[cf. e.g.,][]{donoho_50_2017}.
Using computers and complex software to make sense of (large amounts of) collected or simulated data is a cross-cutting phenomenon leading to _X Data Science_ disciplines, seminars, and study majors.
The practice to apply a combination tools and software stacks for analysing data and the cross-cutting skillsets connected with data science puts the management of computing environments and the challenges of reproducibility high up on the agenda.
This awareness leads to a large number of bespoke Docker images but also some commonly used image stacks, ranging from base images with very common tools to 'catch all' images including _a lot_ of software.
Here we only mention a few image stacks as examples, which are not based on Rocker images or stem from the R community, but prominently include installations of R.
A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).

@@

**Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
It includes machine learling libraries such as Tensorflow and Keras, and also configures the \pkg{reticulate} package.
The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.

@@

As a further example, [**Radiant** project](https://radiant-rstats.github.io/docs/) provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \pkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9GB.

@@

**Gigantum** ([http://gigantum.com/](http://gigantum.com/) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects. 
It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
The independent image stack can be traced back to the Gigantum environment and its features.
The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.} so that appropriate GPU drivers can be enabled automatically when supported.
Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusion}).
Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.

**Whole Tale** (https://wholetale.org) is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole Tale a `tale` is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
For R users, Whole Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
Using Whole Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
Additionally, Whole Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS required to support the Rocker Project images. 
Whole Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability \footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
This works in part because Rocker images are based on a Debian distribution.

@@

Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle. Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\pkg{dockerfiler}), \pkg{containerit}), building images and launching a containers (\pkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
As such, the choice of which package is most useful depends on the use-case at hand, but also the users level of expertise.

```{r clientfunctionalities, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
functionalities <- tribble(
  ~id, ~Functionality,
  "dockerfile", "Generate a Dockerfile",
  "build", "Build an image",
  "execute", "Execute a container locally or remotely",
  "cloud", "Deploy or manage an instances in the cloud",
  "interact", "Interact with an instance (e.g. file transfer)",
  "images", "Manage storage of images",
  "othercontainers", "Supports Docker and Singularity",
  "api", "Direct access to Docker API instead of using the CLI",
  "install", "Installing Docker software"
)

@@

**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is not actively maintained anymore, but should be honorably mentioned as the first R package for managing Docker images and containers.
It uses the \pkg{sys} package to run system commands against the Docker CLI, both locally and through an ssh connection, and has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
The output of container executions are converted to appropriate R types.
The Docker CLI, while evolving quickly and not with great concern for introducing breaking changes, basic functionality is unchanged for a long time so `harbor::docker_run(image = "hello-world")` works like a charm.

@@

**\pkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
It interfaces with the Docker daemon over the Docker HTTP API\footnote{} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
The package is the only one not using system calls to the `docker` CLI for managing images and containers and easily exposes connections to remote Docker daemons, which has to be configured on the Docker level otherwise.
Using the API gives access to more information and is system independent and likely more reliable than parsing command line output.
\pkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/tree/master/inst/spec}{https://github.com/richfitz/stevedore/tree/master/inst/spec} and \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.

**\pkg{AzureContainers}** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), see Section&nbsp;\nameref{azurecontainers}.
While it is mainly intended for working with Azure, as a convenience feature it includes lightweight, cross-platform shells to Docker and Kubernetes (tools `kubectl` and `helm`).
These can be used to create and manage arbitrary Docker images and containers, as well as Kubernetes clusters on any platform or cloud service.

**\pkg{babelwhale}** ([https://cran.r-project.org/package=babelwhale](https://cran.r-project.org/package=babelwhale)) 
allows executing and interacting with containers, which can use either Docker or Singularity as a backend.
The package provides a unified interface to interact with Docker and Singularity containers.
Users can, for example, execute a command inside a container, mount a volume or copy a file.

@@

**\pkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) with Docker: https://rstudio.github.io/renv/articles/docker.html

@@

**\pkg{liftr}** ([https://nanx.me/liftr/](https://nanx.me/liftr/)) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
However, the reproducibility of such content authoring environments is often limited to individual machines --- it is not easy to replicate the system environment (libraries, R versions, R packages) where the document was compiled.
This issue becomes even more serious when it comes to collaborative document authoring and creating large-scale document building services.
_liftr_ solves this reproducibility problem by bringing Docker to the game.
In essence, _liftr_ helps R Markdown users create and manage Docker containers for rendering the documents, thus make the computations utterly reproducible across machines and systems.

@@

Such rApache applications have been used at [rapporter.net](http://rapporter.net), transforming parametrized R Markdown-like documents into interactive web applications for reporting.
A similar design was implemented at [CARD.com](https://www.card.com/), where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and rApache running in Docker containers:
the R session behind Rserve acted as a read-only, exteremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
Similar dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker).
The images extend `rocker/r-base`, e.g., with an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/).
The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution to minimize the risk of breaking the automated builds of the Docker images and production jobs.

A more recent related work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.

@@

Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible and useful \citep{haydel_enhancing_2015}.
GPUs are increasingly popular for compute-intensive machine learning (ML) tasks, e.g., deep artificial neural networks \citep{schmidhuber_deep_2015}.
Though in this case, containers are not completely portable between hardware environments, but the software stack for ML with GPUs is so complex to set up that a ready-to-use container is helpful.
Containers running GPU software require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugin for Docker and a set of base images with appropriate drivers and libraries.
The Rocker&nbsp;Project [has a repository](https://github.com/rocker-org/ml) with (beta) images based on these that include GPU-enabled versions of machine-learning R packages, e.g., `rocker/ml` and `rocker/tensorflow-gpu`.

@@

# Conclusions

This article is a snapshot of the R-corner in a universe of applications built many-faced piece of software, Docker.
`Dockerfile`s and Docker images are the go-to methods for collaboration between roles in an organisation, such as development and IT operations teams, and between parties in the communication of knowledge, such as research workflows or education.
Docker became synonymous with applying the concept of containerisation to solve challenges of reproducible environments, e.g., in research and in development \& production, and of scalable deployments with the ability to move processing between machines easily (e.g., locally, one cloud providers VM, another cloud provider's Container-as-a-Service).
Collaboration, reproducibility, and portability are the common themes behind R packages, use cases, and applications in this work.
